On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-11 Manuscript/manuscript.tex
Overhauled entire document

%1
After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their Youtube
channel that show the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.

@@

%3
In 2009, the International Association of Athletics Federations (Also referred to
as IAAF or World Athletics), which is the governing board for international running
competitions, commissioned a study into changing the reaction time barrier.
The paper analyzed sprinters reaction times and the biological processes 
that went into reacting to the starting gun and concluded 
that athletes can react in as low as 0.08 seconds.  One study by Pain looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}.
to IAAF that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
This could provide a more accurate representation of reaction time, and thus the
possibility of a false disqualification would be lowered \citep{komi2009iaaf}.  
A follow-up study from 2012 by Pilianidis, Mantzouranis, and Kasabalis found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  However, not
all studies have been in favor of a decrease in the reaction time barrier.
Brosnan in 2016 argues for increasing the reaction time barrier to 0.115 seconds,
and cited evidence from European Championships and World Championships in his
article \citep{brosnan2017effects}.  A study looking at the 2008 Olympics
likewise supported the conclusion that athletes could not react in as little as
0.1 seconds \citep{lipps2011implications}.
%About timing
It has been concluded that

%Reaction time barrier has been examined before: cite reviews
%Then go into specifics and talk about people who wanted to change barrier
%General reviews are important
%People have studied the biological processes, cite a few references
%How barrier is picked by regulator
%Read abstract and into and summarize into about one sentence what each paper contributes
%1 and 2 can be merged but 3 need to stand alone

%4
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
When determining the most effective model to use, several variations were
explored across the 4 different data sets: a linear model without a venue effect,
linear model with venue effect, a gamma model without a venue effect, a gamma model
with a venue effect, a gamma model without a venue effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed 
linear models with various distributions to be modeled\citep{Rpkg:lme4}.  The
model we use is a liner model with a fixed effect and random effect and follows
the gamma distribution. The fixed effect comes in the form of an intercept, and the
random effect will be called the venue effect and is reflective of how significant
the changes in the reaction time is for each venue, i.e. the venue effect is 
responsible for the variation between years. 
Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world-class athletes, to not improve significantly from one year to the
next.  In order to reduce any chances of athletes improving reaction time, we
can also look at data from qualifying meets.  Almost every athlete had to qualify 
by proving they are elite at the national level, usually by coming in at least
fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August.  Thus, it should be expected that 
elite athletes such as these will not get reasonably faster in such a short time
span. The equipment used to measure reaction time cannot be guaranteed to be
the same from national to international competition, but the time span between
meets has been reduced considerably.
%Simplify contribution to one paragraph: collect data and formal statistical
%analysis such as model used
%Move a lot of this to methods section

@@

%Follow Anthony's work, explicit references as to what each section does

%Seiko section should go in the Results section.  From results it seems reasonable
%question the data and talk about Seiko

@@

  \begin{equation}
    Y_{ij} = \beta_0 + \beta_1 X_{ij} + \gamma_i + \epsilon_{ij}
  \end{equation}

    where $Y_{ij}$ is the response variable for observation $j$ in year $i$, 
    $X_{ij}$ is the predictor variable for observation $j$ in year $i$, 
    $\beta_0$ is the intercept, $\beta_1$ is the coefficient for $X_{ij}$, 
    $\gamma_i$ is the random effect term for year $i$, and $\epsilon_{ij}$ 
    is the error term for observation $j$ in year $i$.

  %yij denotes ith years and j is observation in year
  %yij will have clustered i and we use random effect to account for this
  %When model is presented talk about how I modeled yij
  %Zij where player is i and cluster, group is year and j
  %Test hypothesis that reaction times are the same vs that they are not
Something that is extremely important to note in the Data is the inconsistency of World Athletics 
in terms of how they penalize false starts.  From 2007-2009, the IAAF (the former name for World 
Athletics, and the governing body for the World Track and Field Champions) instituted a rule change
that allowed one warning false start before a sprinter was disqualified \citep{False-Start}  This 
essentially gave sprinters a second chance and reduced the penalty for a false start.  There were 
18 male and 7 female false starts at the 2007 World Championships, 18 male and 7 female starts at 
the 2009 World Championships. Starting in 2011, the rule was scrapped, and the old policy which 
automatically disqualifies runners who false start was reinstituted.  The effect? At the 2011 World
Championship there were 6 male disqualifications and 4 female disqualifications \citep{False-Start}. 
By returning to the harsher policy and cracking down on false starts, World Athletics had reduced the
number of false starts by two thirds. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{Pilianidis}.

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a box plot as show in Figure~\ref{fig:Boxplots}.  Included are
box plots of the reaction times of the men's semi finals and finals pooled
together and just the finals.  It is clear that the 2022 boxplot for both the
men's finals and semifinals data is lower (indicating a faster reaction time)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
Brosnan found reaction times of 0.156 seconds for men when looking at data from
1999 to 2014 \citep{brosnan2017effects}.
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio

\begin{figure}[h]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \label{fig:Boxplots}
  \caption{The reaction times from 2019 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom)}
\end{figure}

This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

%Throughout exploratory analysis, gamma provides better AIC
%Spend paragraph on after fitting the model how to figure out 0.1 p value

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, or a year
and batch effect.  Density plots were created that simulated the probability of a mean
density less than 0.1, which is the current threshold for a reaction time
disqualification.  The density plots are shown below.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.

%T, generate from fitted model explain what
%The R code does.  Simulate from fitted model. Use distribution of model to 
%Figure out the p value of reaction time less than 0.1
To try and approximate the distribution of the reaction times, R code was written
to try and estimate the probabilities of observing a reaction time as extreme as
0.1 seconds.  The code re-paramertizes the gamma distribution to be in terms of
$\mu$ and shape rather than by $\alpha$ and $\beta$.

At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

\begin{center}
  \begin{tabular}{|c c|} 
   \hline
   Data Set & Probability RxnTime < 0.1 \\ 
   \hline\hline
   all finals gamma log & 0.000864 \\ 
   \hline
   old finals gamma log & 0.000429 \\
   \hline
   all pooled gamma log & 0.006847 \\
   \hline
   old pooled gamma log & 0.005749 \\
  \end{tabular}
  \end{center}

@@

\section{Results}
\jy{summarize your findings; discuss limitations and possible future works.}

%Table should go to the results section
All Men's semi finals and finals data
\begin{center}
  \begin{tabular}{|c  c  c  c  c  c  c |}
   \hline 
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   linear one & -1667 & 835.5 & 0.1529 & 0.0014 & N/A & N/A \\
   \hline
   linear year & -1713 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\ 
   \hline
   gamma one & -1756 & 880.2 & 6.554 & 0.0598 & N/A & N/A \\
   \hline
   gamma year & -1848 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   gamma one log & -1756 & 880.2 & -1.880 & 0.0091 & N/A & N/A \\
   \hline
   gamma year log & -1848 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}
%Section 4.1: Gamma model results show the fitting with and without 2022,
%talk about p values and thresholds from the simulations (do not include the
%results in the methods section). 
Using the gamma function with link set to log (whose summary statistics can be
found in the last row of the above table), the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0425\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.  That means that we would expect a time below the threshold to occur
once every 2,352 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 588 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

It is worth noting that changing the reaction time barrier from 0.1 seconds to
0.08 seconds drastically reduces the chances of observing a reaction time that
would break the barrier.  For the dataset of all finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6.67\cdot10{-7}$.
If IAAF were to change the reaction time the way that \citet{Komi}'s paper
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

@@

%Section 4.2 is motivated by seiko timing and look at same athletes across
%multiple types of competition: 2019 and national

@@

%Wrap up and Conclusion
It is worth noting that in the semi-finals of the World Track and
Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
the legal limit.  What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun.  It seems unlikely that
Allen would be able to correctly predict the gun with such precision two times in a row,
which is the exact reason that the 0.10 second reaction time disqualification barrier was
imposed in the first place.  But Allen showed that at the Championship he was able to
react extremely fast two times, which may suggest skill rather than luck. Devon 
Allen is a very good athlete, as evidenced by his football career at the University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
Thus the 0.099 reaction time may not have been a product of Devon Allen
predicting the start of the race but rather a combination of a quick reaction 
and a possibly faulty sensor.

@@

\section{Appendix}
\label{sec:appendix}
%Put downloadable data spreadsheet
Here is World Athletics website with the data: \url{https://www.worldathletics.org/results/world-athletics-championships}.
Here is the data for the USA Track and Field Championships: \url{https://www.flashresults.com/2022_Meets/Outdoor/06-23_USATF/}

>>>>

%1
After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their Youtube
channel that show the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.
For track and field athletes, they usually only get the chance to prove 
themselves to sponsors every two years, meaning that races such as the finals
of the World Championship is an impactful way to earn a living. An 
investigation into reaction times is also important because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.

@@

%3
In 2009, the International Association of Athletics Federations (Also referred to
as IAAF or World Athletics), which is the governing board for international running
competitions, commissioned a study into changing the reaction time barrier.
The paper analyzed sprinters reaction times and the biological processes 
that went into reacting to the starting gun and concluded 
that athletes can react in as low as 0.08 seconds.  One study by Pain looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. Another study commissioned
by IAAF concluded that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
\citep{komi2009iaaf} This could provide a more accurate representation of 
reaction time, and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.
A 2012 follow-up study to Komi's by Pilianidis found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This contradicts
what was found in a 2021 study that concluded the year of competition has a
significant effect on reaction times \citep{zhang2021correlation}.  Zhang's
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, athletes reaction times got
faster. However, not  all studies have been in favor of a decrease in the
reaction time barrier. Brosnan in 2016 argues for increasing the reaction time 
barrier to 0.115 seconds, and cited evidence from European Championships and 
World Championships in his article \citep{brosnan2017effects}.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
 react in as little as 0.1 seconds \citep{lipps2011implications}.
It is worth noting that one study looking at data from various IAAF meets found
that the fastest reaction times for men were found between the ages of 26 and 29
\citep{tonnessen2013reaction}.  This is notable because Allen was 27 years old
at the 2022 Track and Field World Championship and in his physical peak.

%How barrier is picked by regulator

%4
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a linear model, following a gamma distribution with a random
effect, which will be referred to as the year effect.  The year effect is
reflective of how significant the changes in the reaction time is for each venue,
i.e. the venue effect is  responsible for the variation between years.  
Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who
competed at the World Track and Field Championship in both 2019 and 2022 and
compare the times, and secondly we can look at how athletes performed compared
to how they did at their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

%Simplify contribution to one paragraph: collect data and formal statistical
%analysis such as model used
%Move a lot of this to methods section

@@

@@

Something that is extremely important to note in the Data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007-2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Champions) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{False-Start}  This 
essentially gave sprinters a second chance and reduced the penalty for a false 
start.  There were 18 male and 7 female false starts at the 2007 World 
Championships, 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstituted.  The effect? At the 2011 
World Championship there were 6 male disqualifications and 4 female 
disqualifications \citep{False-Start}. By returning to the harsher policy and 
cracking down on false starts, World Athletics had reduced the number of 
false starts by two thirds. One study that looked at different false start
rules that IAAF has imposed from 1997 to 2009 and found that the reaction 
times when there was less leniency improved by a statistically significant amount
\citep{haugen2013effect}. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{Pilianidis}.  However, we chose not to include women's data in this
paper due to the numerous papers citing difference in reaction times in men and
women's reaction times: \citet{lipps2011implications}, \citet{babicc2009reaction}
\citet{panoutsakopoulos2020gender}.

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the
appendix~\ref{sec:Appendix} 
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical significance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). Included are
box plots of the reaction times of the men's finals and the men's semifinals and
finals pooled together.  We will refer to the latter as the "pooled data"
throughout the paper.  The heats or preliminary rounds were not studied because
athletes react faster as the competition progresses.  Athletes react faster in
the finals than the semifinals which are faster than the heats 
\ref{zhang2021correlation}.  If we want to determine if Allen's reaction time
is reasonable, then we thus need to look at the men's final data, with the men's
pooled data being included in order to increase the sample size.  In 2022, there
were only five data points due to two disqualifications and one athlete who 
qualified but did not compete.
The data that looked at in the paper is best visualized using a box plot as show 
in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both the
men's finals and semifinals data is lower (indicating a faster reaction time)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
Brosnan found reaction times of 0.156 seconds for men when looking at data from
1999 to 2014 \citep{brosnan2017effects}.
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio

\begin{figure}[tbp]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \label{fig:Boxplots}
  \caption{The reaction times from 2019 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom)}
\end{figure}

This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

@@

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, 
or a year and batch effect.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.
Thus, we decided to inclue only the year effect in the GLMM, which is presented
as:

\begin{equation}
  \label{eq:GLMM}
  Y_{ij} = \beta_0 + \gamma_i + \epsilon_{ij}
\end{equation}

Where $Y_{ij}$ is the response variable for observation $j$ in year $i$, , 
$\beta_0$ is the intercept, $\gamma_i$ is the random effect term for year $i$, 
and $\epsilon_{ij}$ is the error term for observation $j$ in year $i$.

%yij denotes ith years and j is observation in year
%yij will have clustered i and we use random effect to account for this
%When model is presented talk about how I modeled yij
%Zij where player is i and cluster, group is year and j
%Test hypothesis that reaction times are the same vs that they are not

%Throughout exploratory analysis, gamma provides better AIC
%Spend paragraph on after fitting the model how to figure out 0.1 p value
As the GLMM model has mixed form, the best way to try to gauge the probability
of observing a reaction time is to simulate the distribution of times using a 
large sample.  In order to approximate the distribution, R code was written to 
try and estimate the probabilities of observing a reaction time as extreme as
0.1 seconds.  The code re-paramertizes the gamma distribution to be in terms of
$\mu$ and shape rather than by $\alpha$ and $\beta$.  From the code, it is
possible to calculate the probability of observing a reaction time faster than
a given time.  We want the probability of a reaction time being less than 0.1
seconds in order to gauge if that is a reasonable disqualification barrier.

%T, generate from fitted model explain what
%The R code does.  Simulate from fitted model. Use distribution of model to 
%Figure out the p value of reaction time less than 0.1

@@

\section{Results} \label{sec:Results}
\jy{summarize your findings; discuss limitations and possible future works.}

\subsection{GLMM Results}
%Section 4.1: Gamma model results show the fitting with and without 2022,
%talk about p values and thresholds from the simulations (do not include the
%results in the methods section). 
At a size of n=10,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

\begin{center}
  \begin{tabular}{|c c|} 
   \hline
   Data Set & Probability RxnTime < 0.1 \\ 
   \hline\hline
   all finals gamma log & 0.000856 \\ 
   \hline
   old finals gamma log & 0.000439 \\
   \hline
   all pooled gamma log & 0.006865 \\
   \hline
   old pooled gamma log & 0.005761 \\
  \end{tabular}
  \end{center}

Using the gamma function with link set to log (whose summary statistics can be
found in the last row of the above table), the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.  That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 569 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

It is worth noting that changing the reaction time barrier from 0.1 seconds to
0.08 seconds drastically reduces the chances of observing a reaction time that
would break the barrier.  For the dataset of all finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6\cdot10{-7}$.
If IAAF were to change the reaction time the way that \citet{Komi}'s paper
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

@@

\subsection{Rank Comparison Results}
%Section 4.2 is motivated by seiko timing and look at same athletes across
%multiple types of competition: 2019 and national

@@

\section{Discussion}\label{sec:Discussion}

One conclusion from \citet{zhang2021correlation}'s analysis was that athletes
reaction times improved from 2011 to 2019.  However this study does not consider 
data from 1999 to 2009, and over this timespan athletes get slower and then
faster.  A boxplot~\ref{fig:Boxplot} shows that for the men's pooled data, the 
year with the fastest median reaction time before 2022 was 1999.  Thus, this 
reject's Zhang's conclusion that athletes are getting faster because they were 
fast over twenty years ago.  Instead, a more appropriate conclusion may be that
there is a lot of variability in reaction times as evidenced by the two fastest
years occurring 23 years apart.

It is worth noting that in the semi-finals of the World Track and
Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
the legal limit.  What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun.  It seems unlikely that
Allen would be able to correctly predict the gun with such precision two times in a row,
which is the exact reason that the 0.10 second reaction time disqualification barrier was
imposed in the first place.  But Allen showed that at the Championship he was able to
react extremely fast two times, which may suggest skill rather than luck. Devon 
Allen is a very good athlete, as evidenced by his football career at the University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
Thus the 0.099 reaction time may not have been a product of Devon Allen
predicting the start of the race but rather a combination of a quick reaction 
and a possibly faulty sensor.

@@

\section{Appendix}
\label{sec:Appendix}
%Put downloadable data spreadsheet
Here is World Athletics website with the data: \url{https://www.worldathletics.org/results/world-athletics-championships}.
Here is the data for the USA Track and Field Championships: \url{https://www.flashresults.com/2022_Meets/Outdoor/06-23_USATF/}

For the pooled data set
\begin{center}
  \begin{tabular}{|c  c  c  c  c  c  c |}
   \hline 
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   linear one & -1667 & 835.5 & 0.1529 & 0.0014 & N/A & N/A \\
   \hline
   linear year & -1713 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\ 
   \hline
   gamma one & -1756 & 880.2 & 6.554 & 0.0598 & N/A & N/A \\
   \hline
   gamma year & -1848 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   gamma one log & -1756 & 880.2 & -1.880 & 0.0091 & N/A & N/A \\
   \hline
   gamma year log & -1848 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}
